---
title: "Data_605_Final"
author: "Coco Donovan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Computational Mathematics

$\\$
$\\$

Your final is due by the end of the last week of class.  You should post your 
solutions to your GitHub account or RPubs.  You are also expected to make a short
presentation via YouTube  and post that recording to the board.  This project will
show off your ability to understand the elements of the class. 

### Problem 1.

Using R, set a random seed equal to 1234 (i.e., set.seed(1234)).  Generate a random
variable X that has 10,000 continuous random uniform values between 5 and 15.
Then generate a random variable Y that has 10,000 random normal values with a mean
of 10 and a standard deviation of 2.89.

$\\$
$\\$

$Probability$

$\\$

Calculate as a minimum the below probabilities a through c.  Assume the small letter
"x" is estimated as the median of the X variable, and the small letter "y" is 
estimated as the median of the Y variable.  

```{r}
set.seed(1234)

X <- runif(n = 10000, min = 5, max = 15)

Y <- rnorm(n = 10000, mean = 10, sd = 2.89) 

x = median(X)

y = quantile(Y, 0.25)
```

#### Interpret the meaning of all probabilities.

##### a.   $P(X>x | X>y)$
$\\$
$\\$
$P(X>x | X>y) = \frac{P(X>x \cap X>y)}{P(X>y)}$

```{r}
num <- length(X[X>x & X>y])/length(X)
denom <- length(X[X>y])/length(X)

num/denom
```

##### b.  $P(X>x \:\&\: Y>y)$
$\\$
$\\$

```{r}
length(which(X>x & Y>y))/10000
```

##### c.  $P(X<x | X>y)$	
$\\$
$\\$
$P(X>x | X>y) = \frac{P(X>x \cap X>y)}{P(X>y)}$

```{r}
num <- length(X[X<x & X>y])/length(X)
denom <- length(X[X>y])/length(X)

num/denom
```

Investigate whether P(X>x & Y>y)=P(X>x)P(Y>y) by building a table and evaluating
the marginal and joint probabilities.

```{r}
library(magrittr)
library(kableExtra)
library(dplyr)

A <- length(which(X>x & Y>y))/10000
B <- length(which(X<=x & Y>y))/10000
C <- length(which(X>x & Y<=y))/10000
D <- length(which(X<=x & Y<=y))/10000

df <- data.frame(c(A,C),c(B,D)) %>%
  mutate(Total = rowSums(.)) %>%
  rbind(., colSums(.)) %>%
  set_colnames(c('$X>x$', '$X\\leq x$', 'Total')) %>%
  set_rownames(c('$Y>y$', '$Y\\leq y$', 'Total')) %>%
  kable() %>% 
  kable_styling(latex_options = "hold_position")

df
```

From the table, the joint probability is 0.3738 and the marginal probability is 
0.375 (the result of multiplying P(X>x) =0.5 and P(Y>y) = 0.75). While the two 
probabilities are close in value, they are not exactly the same.
$\\$
$\\$

Check to see if independence holds by using Fisher’s Exact Test and the Chi Square
Test.  What is the difference between the two? Which is most appropriate?  Are you
surprised at the results?  Why or why not?
Null Hypothesis: $P(X>x)$ and $P(Y>y)$ are independent
$\\$
Alternative Hypothesis: $P(X>x)$ and $P(Y>y)$ are dependent

```{r}
table(X > x, Y > y) %>%
  fisher.test()
```

The fishers test yields a p-value of 0.5953, which is significantly higher than 
our goal of 0.05 and for this reason we cannot reject the null hypothesis based
on the Fisher's exact test.

```{r}
table(X>x, Y>y) %>%
  chisq.test(.)
```

The Chi-squared test here yields the same p-value as the Fisher's Exact test, meaning
that based on the Chi-squared test (just like the Fisher's Exact Test), we cannot
reject the Null hypothesis ($P(X>x)$ and $P(Y>y)$ are independent).

$\\$
$\\$

Fisher's Exact test is best used on a smaller sample size whereas Chi-squared 
produces better results as the sample size. With that said, it would make sense 
to use Chi-squared in this case; however, this may not be the best illustration
because the p-values ended up being the same here.

### Problem 2

$\\$
$\\$

You are to register for Kaggle.com (free) and compete in the Regression with a Crab Age Dataset competition.  https://www.kaggle.com/competitions/playground-series-s3e16  I want you to do the following.

$\\$
$\\$ 

Here I load the training data: 

```{r}
library(readr)
crabs <- read_csv('train.csv')
```
 
$Descriptive \: and \: Inferential \: Statistics$

$\\$

Provide univariate descriptive statistics and appropriate plots for the training data set.  Provide a scatterplot matrix for at least two of the independent variables and the dependent variable. Derive a correlation matrix for any three quantitative variables in the dataset.  Test the hypotheses that the correlations between each pairwise set of variables is 0 and provide an 80% confidence interval.  Discuss the meaning of your analysis.  Would you be worried about familywise error? Why or why not?

###### Summary:

```{r}
summary(crabs)
```

$\\$
$\\$
```{r}
pairs(~`Shucked Weight`+Weight+Height+`Shell Weight`, data = crabs, lwd = 0.5)
```

```{r}
vars <- c("Weight", "Height", "Shucked Weight")

corr_matrix <- cor(crabs[vars])

corr_matrix
```

Now, I'll go ahead and check the pairwise correlation of Weight, Height and 
Schucked Weight.

```{r}
cor.test(~ Weight + Height, data = crabs, method = "pearson", conf.level = 0.8)

cor.test(~ `Shucked Weight` + Height, data = crabs, method = "pearson", conf.level = 0.8)

cor.test(~ Weight + `Shucked Weight`, data = crabs, method = "pearson", conf.level = 0.8)
```

All three variable combinations yield a p-value far less than 0.05 and none of the 
confidence intervals contain 0. These results tell us that there is likely a meaningful
relationship to explore between these variable pairing relationships.

Familywise error involves a type one error or rejecting a null hypothesis that is
actually true in the population. I would not be worried about that being true in
this scenario because the p-value is almost 0.

$Linear \: Algebra \: and \: Correlation$

$\\$

Invert your correlation matrix from above. (This is known as the precision matrix and contains variance inflation factors on the diagonal.) Multiply the correlation matrix by the precision matrix, and then multiply the precision matrix by the correlation matrix. Conduct LDU decomposition on the matrix. 

$\\$
$\\$

```{r}
# here I invert the correlation matrix
inverted_mat <- solve(corr_matrix)

inverted_mat

# here I multiply the inverted matrix by the correlation matrix
pre_cor <- inverted_mat %*% corr_matrix

pre_cor

# here I multiply the correlation matrix by the inverted matrix
cor_pre <- corr_matrix %*% inverted_mat

cor_pre
```

Now, I'll conduct an LU Decomp for the initial correlation matrix:

```{r}
library(matrixcalc)

lu.decomposition(corr_matrix)
```

$Calculus-Based \: Probability \: \& \: Statistics$

$\\$

Many times, it makes sense to fit a closed form distribution to data.  Select a variable in the Kaggle.com training dataset that is skewed to the right, shift it so that the minimum value is absolutely above zero if necessary.  Then load the MASS package and run fitdistr to fit an exponential probability density function.  (See  https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/fitdistr.html ).  Find the optimal value of λ for this distribution, and then take 1000 samples from this exponential distribution using this value (e.g., rexp(1000, λ)).  Plot a histogram and compare it with a histogram of your original variable.   Using the exponential pdf, find the 5th and 95th percentiles using the cumulative distribution function (CDF).   Also generate a 95% confidence interval from the empirical data, assuming normality.  Finally, provide the empirical 5th percentile and 95th percentile of the data.  Discuss.

$\\$
$\\$

Here, I'll inspect for right-skewness:

```{r}
library(ggplot2)
library(tidyr) # so I can have access to gather ()

# subsetting to only numeric values

crabs_numeric <- subset(crabs, select = -c(id, Sex))

# preliminary looks for skewness

ggplot(gather(crabs_numeric), aes(value)) + 
    geom_histogram(bins = 10) + 
    facet_wrap(~key, scales = 'free_x')
```

```{r}
library(e1071)

skews <- crabs_numeric %>%
  summarise(across(where(is.numeric), ~ skewness(.x, na.rm = TRUE)))

min(crabs$Age) > 0
```

Age has a skew of about 1.093, which means it is skewed heavily to the right and 
the minimum value of crabs\$Age is greater than zero, which means that shifting 
the variable (in this case) would not be needed.

```{r}
library(MASS)

exp_age <- fitdistr(crabs$Age, "exponential")

lambda_age <- exp_age$estimate

exp_sample <- rexp(1000, lambda_age)

summary(exp_sample)
```

##### Histogram:

```{r}
hist(exp_sample, main = "Histogram of Exponential Sample of Crabs Age")
```



```{r}
lower <- qexp(0.05, lambda_age)
upper <- qexp(0.95, lambda_age)

print(paste('5th percentile: ', lower))
print(paste('95th percentile: ', upper))
```



```{r}
quantile(crabs$Age, c(0.05, 0.95))
```

$Modeling$
$\\$
$\\$

```{r}
# No missing values
colSums(is.na(crabs_numeric))
```



```{r}
crabs_numeric <- subset(crabs, select = -c(id, Sex))

crabs_model <- lm(Age ~., crabs_numeric)
summary(crabs_model)
```

All variables have an associated p-value which is less than 0.05 and an associated 
t-value with is more than 1.96 away from 0 meaning they are likely statistically 
significant. The R-squared values are both a little below .54 meaning that the model
can account for about 54% of variation.

```{r}
plot(crabs_model)
```



```{r}
unique(crabs$Sex)

crabs <- crabs %>%
  group_by(Sex) %>%
  mutate(mean_weight = mean(Weight))

crabs_numeric <- subset(crabs, select = -c(id, Sex))

crab_model2 <- lm(Age ~ ., data = crabs_numeric)
summary(crab_model2)
```

Initially had made a really complicated model and that yielded an R-Squared of
to about 0.57; however the p-values of some variables grew and the t-values shrunk.

The variables for my new model all have an associated p-value which is less than 
0.05 and an associated t-value with is more than 1.96 away from 0 meaning they 
are likely statistically significant. Though, in this new model, I was able to 
improve the R-squared values to about .55, meaning this new model can account for 
more variation.

```{r}
plot(crab_model2)
```

And if you look at the plots of my residuals, the residuals seem to be more evenly 
dispersed above and below zero, not to mention that the QQ-plot displays residuals
that more closely adhere to the normal line than do the plots from the first model.

```{r}
crab_test <- read.csv('test.csv')

crab_test <- crab_test %>%
  group_by(Sex) %>%
  mutate(mean_weight = mean(Weight))

crab_test <- crab_test %>%
  rename(`Shucked Weight`=Shucked.Weight, `Viscera Weight`=Viscera.Weight, `Shell Weight`=Shell.Weight)

prediction <- predict(crab_model2, crab_test, type = "response")
head(prediction)

kaggle_predictions <- data.frame(Id = crab_test$id, Age = prediction)
head(kaggle_predictions)


# I just needed to run this once, but I do not want this to run everytime I knit
write.csv(kaggle_predictions, file = "crab_age2.csv", row.names=FALSE)
```

It say the deadline for submission has past and therefore I cannot submit my predictions
to the leaderboard. However, my score was a 1.48.
